Dependent Object Types

### 2012-08-05 article.tex
original dot-one

%%% BEGIN RULES %%%

@@

@@

%%% END RULES %%%

>>>>

@@

\end{multicols}
} % END dot-one

The DOT calculus is a simple system of dependent
object-types. Figure~\ref{dot-one} gives its syntax, reduction rules,
and type assignment rules.

\subsection*{Notation} We use standard notational conventions for
sets. The notation $\seq{X}$ denotes a set of elements $X$. Given a
such a set $\seq X$ in a typing rule, $X_i$ denotes an arbitrary
element of $X$. 
The $\uplus$ operator extends a set of bindings. It is required that the added binding does not
introduce a variable which is already bound in the base-set. 
We use an
abbreviation for preconditions in typing judgements. Given an
environment $\Gamma$ and some predicates $P$ and $Q$, the condition $\Gamma \ts P \spcomma Q$
is a shorthand for the two conditions $\Gamma \ts P$ and $\Gamma \ts Q$.

\subsection*{Syntax}

There are three alphabets: Variable names $x$, $y$, $z$ are freely
alpha-renamable. They occur as parameters of lambda abstractions, as
binders for objects created by \verb@new@-expressions, and as self
references in refinements. Value labels $l$ denote fields in objects,
which are bound to values at run-time. Type labels $L$ denote type
members of objects. Type labels are further separated into labels for
abstract types $L_a$ and labels for classes $L_c$. It is assumed that
in each program every class label $L_c$ is declared at most once.

We assume that the label alphabets $l$ and $L$ are finite. This is
not a restriction in practice, because one can include in these 
alphabets every label occurring in a given program.

The terms $t$ in DOT consist of variables $x$, $y$, $z$, lambda
abstractions $\abs x T t$, function applications $\app t {t'}$, field
selections $t.l$, and object creation expressions $\new x c t$ where
$c$ is a constructor $T_c \ldefs{\seq{l = v}}$. The latter binds a
variable $x$ to a new instance of type $T_c$ with fields $\seq l$ initialized to values $\seq v$.  The scope of $x$ extends through the term $t$.

Two subclasses of terms are values $v$, which consist of just
variables and lambda abstractions, and paths $v$ which consist of just
variables and selections.

The types in DOT are denoted by letters $S$, $T$, $U$, or $V$. They consist of the following:
\begin{itemize}
\item[-] Type selections $p.L$, which denote the type member $L$ of path $p$.
%         A shorthand notation applies to type selections on singleton types: Instead
%         of $p \sing \tsel L$ one can write just $p.L$.
\item[-] Refinement types $T \refine z {\seq D}$, which refine a type $T$ by a set of declarations $D$.
         The variable $z$ refers to the ``self''-reference of the type. Declarations can refer to
         other declarations in the same type by selecting from $z$.
\ifdep
  \item[-] Dependent function types $\dfun x T {T'}$.
\else
  \item[-] Function types $T \tfun T'$.
\fi
\item[-] Type intersections $T \tand T'$, which carry the declarations of members present in either $T$ or $T'$.
\item[-] Type unions $T \tor T'$, which carry only the declarations of members present in both $T$ and $T'$.
\item[-] A top type $\Top$, which corresponds to an empty object.
\item[-] A bottom type $\Bot$, which represents a non-terminating computation.
\end{itemize}
A subset of types $T_c$ are called {\em concrete types}. These are type selections
$p.L_c$ of class labels,
the top type $\Top$, intersections of concrete types, and refinements $T_c \refine z {\seq D}$ of concrete types. Only concrete types are allowed in constructors $c$.

%Types in DOT form a lattice where $\tand$ is join and $\tor$ is meet.

There are only two forms of declarations in DOT, which are both part
of refinement types.  A value declaration $\ldecl l T$ introduces a
field with type $T$.  A type declaration $\Ldecl L S U$ introduces a
type member $L$ with a lower bound type $S$ and an upper bound type
$U$. There are no type aliases, but a type alias can be simulated by a
type declaration $\Ldecl L T T$ where the lower bound and the upper
bound are the same type $T$.

%It is assumed that the lower bound of
%every class type is the bottom type and that the upper bound is a
%concrete type. That is, the only permissible form of a class
%declaration is $\Ldecl {L_c} \Bot T_c$.

Every field or type label can be declared only once in a set of
declarations $\seq D$. A set of declarations can hence be seen as a map from
labels to their declarations.  Meets $\wedge$ and joins $\vee$ on sets of
declarations are defined as follows.

      \dom(\seq D \tand \seq {D'}) &~=~& \dom(\seq{D}) \cup \dom(\seq{D'}) \\
      \dom(\seq D \tor \seq {D'}) &=& \dom(\seq{D}) \cap \dom(\seq{D'}) \\[0.5em]
      (D \tand D')(L) &=&
        \Ldecl L {(S \tor S')} {(U \tand U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
        &=& D(L) & \mbox{if~} L \notin \dom(\seq{D'}) \\
        &=& D'(L) & \mbox{if~} L \notin \dom(\seq{D}) \\
      (D \tand D')(l) &=&
        \ldecl l {T \tand T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} \\
        &=& D(l) & \mbox{if~} l \notin \dom(\seq{D'}) \\
        &=& D'(l) & \mbox{if~} l \notin \dom(\seq{D}) \\[0.5em]
      (D \tor D')(L) &=&
        \Ldecl L {(S \tand S')} {(U \tor U')} & \mbox{if~} (\Ldecl L S U) \in \seq{D} \;\mbox{and}\; (\Ldecl L {S'} {U'}) \in \seq{D'} \\
      (D \tor D')(l) &=&
        \ldecl l {T \tor T'} & \mbox{if~} (\ldecl l T) \in \seq{D} \;\mbox{and}\; (\ldecl l {T'}) \in \seq{D'} 
    \eda

Sets of declarations form a lattice with the given meet $\wedge$ and
join $\vee$, the empty set of declatations as the top element, and
the bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
declarations that contains for every term label $l$ the
declaration $\ldecl l \Bot$ and for every type label $L$ the
declaration $\Ldecl L \Top \Bot$.

\subsection*{Reduction rules}

Reduction rules $\reduction t s {t'} {s'}$ in DOT rewrite pairs of
terms $t$ and stores $s$, where stores map variables to constructors.
There are three main reduction rules: ($\beta_V$) is the standard beta
reduction of call-by-value lambda calculus. Rule (\textsc{new}) rewrites an
object creation $\new x c t$ by placing the binding of variable $x$ 
to constructor $c$ in the store and continuing with term $t$. Rule (\textsc{sel})
rewrites a field selection $x.l$ by retrieving the corresponding value from the store.
These reduction rules can be applied anywhere in a term where the hole $[\,]$ of 
an evaluation context $e$ can be situated.

\subsection*{Type assignment rules}

The last part of Figure~\ref{dot-one} presents rules for type
assignment.  
\ifdep
The four rules (\textsc{var}), (\textsc{abs}), (\textsc{app-term}) and
(\textsc{app-path}) correspond to dependently typed lambda calculus
with a subtyping discipline. There are two rules for
applications. Rule (\textsc{app-term}) implements the standard
application rule if the function's result type does not depend on its
argument type. Rule (\textsc{app-path}) applies to dependently typed
functions which can only be applied to paths.
Instead of adding a separate subsumption rule,
subtyping is expressed by preconditions in the application rules and
also in rule (\textsc{new}).  
\else
The three rules (\textsc{var}), (\textsc{abs}), and
(\textsc{app}) on the left correspond to simply typed lambda calculus
with a subtyping discipline (I believe it is straightforward to extend
this to full $F_{\sub}$). Rule (\textsc{abs}) has an additional precondition
stating that the bound variable $x$ may not appear in the function's
result type $T'$. It would be an interesting extension of the calculus
to drop this precondition, thereby allowing dependent function types.
Instead of adding a separate subsumption rule,
subtyping is expressed by preconditions in rules (\textsc{app}) and
(\textsc{new}).  
\fi
Rule rule (\textsc{sub}) is the standard subsumption rule for subtyping.

%The (\textsc{path}) rule gives the alternate type $p \sing$ to a well-typed path
%$p$ whose type $T$ has a non-empty expansion $D,\seq{D}$.  By the subtyping rule
%(\textsc{single-$\sub$}) in Figure~\ref{dot-sub}, the alternate type $p \sing$
%is guaranteed to be a subtype of the underlying type $T$ of $p$.

Rule (\textsc{sel}) types a field selection by means of an auxiliary
membership relation $\ni$, which determines whether a given term contains a given
declaration as one of its members. The membership relation is defined in
Figure~\ref{dot-two} and is further explained below.

The last rule, (\textsc{new}), assigns types to object creation
expressions. It is the most complex of DOT's typing rules.  To
type-check an object creation $\new x {T_c \ldefs {\seq{l = v}}} t$,
one verifies first that the type $T_c$ is well-formed (see
Figure~\ref{dot-wf} for a definition of well-formedness).  One then
determines the set of all declarations that this type carries, using
the expansion relation $\expand$ defined in Figure~\ref{dot-two}.
Every type declaration $\Ldecl L S U$ in this set must be realizable,
i.e.\ its lower bound $S$ must be a subtype of its upper bound $U$.
Every field declaration $\ldecl l V$ in this set must have a
corresponding initializing value of $v$ of type $V$.  These checks are
made in an environment which is extended by the binding $x: T_c$. In
particular this allows field values that recurse on ``self'' by
referring to the bound variable $x$. As in the (\textsc{abs}) rule, it
is required that the bound variable does not appear in the
expression's result type $T'$.

@@
